---
title: "Confirm SCPCP000023 sample overlap with OsteoCar"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
---

Samples from `SCPCP000023` are included in the `OsteoCar` reference.
This exploratory notebook confirms/assesses the extent of sample overlap to determine how to proceed.

## Setup

```{r, warning = FALSE}
# settings
options(
  dplyr.summarise.inform = FALSE, 
  readr.show_col_types = FALSE
)
```


### Paths

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

module_dir <- file.path(repository_base, "analyses", "cell-type-osteosarcoma")
ref_dir <- file.path(module_dir, "references")
data_dir <- file.path(repository_base, "data", "current", "SCPCP000023")
```

## Analysis

First, we'll read in the sample metadata for `SCPCP000023`.

```{r}
dat <- readr::read_tsv(file.path(data_dir, "single_cell_metadata.tsv"))

nrow(dat)
```

These samples comprise both primary and metastatic samples:

```{r}
table(dat$primary_or_metastasis)
```


Next, we'll read in the `OsteoCar` sample details, specifically grabbing the samples from St. Jude.
For this, we'll read in the actual references themselves to compare sample ids.

```{r}
mets_object <- qs2::qs_read(file.path(ref_dir, "patient_mets.qs2"))
met_ids <- mets_object@meta.data |>
  dplyr::distinct(sample_name) |>
  dplyr::filter(stringr::str_starts(sample_name, "SJOS")) |>
  dplyr::pull(sample_name)



prim_object <- qs2::qs_read(file.path(ref_dir, "patient_prim.qs2"))
prim_ids <- prim_object@meta.data |>
  dplyr::distinct(sample_name) |>
  dplyr::filter(stringr::str_starts(sample_name, "SJOS")) |>
  dplyr::pull(sample_name)
```


How many are there?

```{r}
met_ids
```


```{r}
prim_ids
```

Let's see how many of our samples are present in the reference.

First, metastatic samples:

```{r}
intersect(
  dat$submitter_id, 
  met_ids
)
```

All of the metastatic samples are present.

Next, primary samples:


```{r}
intersect(
  dat$submitter_id, 
  prim_ids
)
```
Of the 12 St. Jude ids in the `prim_met`, 11 are in `SCPCP000023`.


To confirm, how many `SCPCP000023` ids are _not_ `OsteoCar`?

```{r}
setdiff(
  dat$submitter_id,
  c(met_ids, prim_ids)
)
```

It looks like we probably have a xenograft (`SJOS013768_X1`) of a sample that is in present in `prim_ids` (`SJOS013768_D1_repB`), so this is the one difference for `prim_ids`.

Ultimately, this means most samples in `SCPCP000023` are _not_ represented in `OsteoCar`, so we should indeed include this project.
We will determine how to handle the samples that also exist in the reference a bit later in this module's development, specifically by comparing their annotations for those samples to ours.
This will be useful for helping us gauge the performance of our annotation approach more generally.

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```
