---
title: "Confirm SCPCP000023 sample overlap with OsteoCar"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
---

Samples from `SCPCP000023` are included in the `OsteoCar` reference.
This exploratory notebook confirms/assesses the extent of sample overlap to determine how to proceed.

## Setup

```{r, warning = FALSE}
# settings
options(
  dplyr.summarise.inform = FALSE, 
  readr.show_col_types = FALSE
)
```


### Paths

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

module_dir <- file.path(repository_base, "analyses", "cell-type-osteosarcoma")
ref_dir <- file.path(module_dir, "references")
data_dir <- file.path(repository_base, "data", "current", "SCPCP000023")
```

```{r}
output_file <- file.path(ref_dir, "SCPCP000023-osteocar-samples.tsv")
```


## Analysis

First, we'll read in the sample metadata for `SCPCP000023`.

```{r}
dat <- readr::read_tsv(file.path(data_dir, "single_cell_metadata.tsv"))

nrow(dat)
```

These samples comprise both primary and metastatic samples:

```{r}
table(dat$primary_or_metastasis)
```


Next, we'll read in the `OsteoCar` sample details, specifically grabbing the samples from St. Jude.
For this, we'll read in the actual references themselves to compare sample ids.

```{r}
# define helper function for grabbing St Jude ids
get_sj_ids <- function(filename) {
  qs2::qs_read(file.path(ref_dir, filename))@meta.data |>
    dplyr::distinct(sample_name) |>
    dplyr::filter(stringr::str_starts(sample_name, "SJOS")) |>
    dplyr::pull(sample_name)
}

osteocar_ids <- c(
  get_sj_ids("patient_mets.qs2"),
  get_sj_ids("patient_prim.qs2"),
  get_sj_ids("xeno_mets.qs2"),
  get_sj_ids("xeno_prim.qs2")
)
```


How many are there?

```{r}
length(osteocar_ids)
```

Let's see how many of our samples are present in the reference.


```{r}
shared_ids <- intersect(
  dat$submitter_id, 
  osteocar_ids
)
shared_ids
```

ScPCA contains 37/38 of these samples.

To confirm, how many `SCPCP000023` ids are _not_ `OsteoCar`?

```{r}
setdiff(
  dat$submitter_id,
  osteocar_ids
)
```

This means most samples in `SCPCP000023` are represented in `OsteoCar`.
We will determine how to handle the samples that also exist in the reference a bit later in this module's development, specifically by comparing their annotations for those samples to ours.
This will be useful for helping us gauge the performance of our annotation approach more generally.

To conclude, we'll export the IDs that are overlapping which will come in handy later when we run this evaluation.


```{r}
dat |> 
  dplyr::filter(submitter_id %in% shared_ids) |>
  dplyr::select(contains("scpca"), submitter_id) |>
  readr::write_tsv(output_file)
```

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```
