---
title: "Compare OsteoCar references"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
---

There are several different references in `OsteoCar`, and while we can generally map them to our sample propertoes, we have some sample which don't neatly line up.
For example, `OsteoCar` has patient primary and metastatic (lung only) references, but we have a sample from metastatic tissue in bone which differs from the metastatic reference. 

The goal of this notebook is to lightly compare some of the references using `SingleR`:
* patient primary vs metastatic (lung only) references
* xenograft primary vs metastatic references
As these references are quite large, they have to be aggregated before use.

We compare the references to one another to see how similar the cell type distributions and expression signatures are.

## Setup

```{r}
suppressPackageStartupMessages({
  library(SingleR)
  library(SingleCellExperiment)
})
bp_param <- BiocParallel::MulticoreParam(6)
set.seed(2026)
```

```{r}
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

module_dir <- file.path(repository_base, "analyses", "cell-type-osteosarcoma")
ref_dir <- file.path(module_dir, "references")
```



## Analysis

We'll compare `Ann_Level2` between primary and metastatic references by first aggregating them and then running `SingleR::matchReferences()` and plotting the resulting heatmap.
This will tell us where cell types overlap/differ between references.

Note that we read in references when we use them and remove them promptly after aggregation, to save memory.

### Primary references

```{r}
patient_prim <- readr::read_rds(file.path(ref_dir, "patient_prim_sce.rds"))
patient_prim_aggr <- SingleR::trainSingleR(
  ref = patient_prim,
  labels = patient_prim$Ann_Level2,
  de.method = "wilcox",
  aggr.ref = TRUE,
  BPPARAM = bp_param
)
rm(patient_prim)

patient_mets <- readr::read_rds(file.path(ref_dir, "patient_mets_sce.rds"))
patient_mets_aggr <- SingleR::trainSingleR(
  ref = patient_mets,
  labels = patient_mets$Ann_Level2,
  de.method = "wilcox",
  aggr.ref = TRUE,
  BPPARAM = bp_param
)
rm(patient_mets)

gc()
```


In the heatmap below, the primary reference is columns and the metastasis reference is rows.

```{r}
matched <- matchReferences(
  patient_mets_aggr$ref, 
  patient_prim_aggr$ref, 
  patient_mets_aggr$labels$full, 
  patient_prim_aggr$labels$full 
)
pheatmap::pheatmap(
  matched, 
  col=viridis::magma(500), 
  cluster_rows = FALSE, 
  cluster_cols = FALSE
)
```

We see that many of the immune cell types as well as endothelial cells match up between references, although with some noise for some of the cell types.
But, a substantial proportion of cell types don't have a clear match in the other reference which tells us that these references generally have distinct profiles.



### Xenograft references

Note that there are many fewer cell types here, because the reference considers only human-derived tumor cells.
Also note, `Ann_Level2` and `Ann_Level3` are the same for the xenograft references, so this is the most fine-grained we can achieve.

```{r}
xeno_prim <- readr::read_rds(file.path(ref_dir, "xeno_prim_sce.rds"))
xeno_prim_aggr <- SingleR::trainSingleR(
  ref = xeno_prim,
  labels = xeno_prim$Ann_Level2,
  de.method = "wilcox",
  aggr.ref = TRUE,
  BPPARAM = bp_param
)
rm(xeno_prim)

xeno_mets <- readr::read_rds(file.path(ref_dir, "xeno_mets_sce.rds"))
xeno_mets_aggr <- SingleR::trainSingleR(
  ref = xeno_mets,
  labels = xeno_mets$Ann_Level2,
  de.method = "wilcox",
  aggr.ref = TRUE,
  BPPARAM = bp_param
)
rm(xeno_mets)

gc()
```


In the heatmap below, the primary reference is columns and the metastasis reference is rows.

```{r}
matched <- matchReferences(
  xeno_mets_aggr$ref, 
  xeno_prim_aggr$ref, 
  xeno_mets_aggr$labels$full, 
  xeno_prim_aggr$labels$full 
)
pheatmap::pheatmap(
  matched, 
  col=viridis::magma(500), 
  cluster_rows = FALSE, 
  cluster_cols = FALSE
)
```

There's a 1:1 match of proliferative here only, but the noise around other cell types suggests that there are distinct expression profiles between shared cell types across these references.
What is not clear is how much of that is signal vs noise. 
The `Interactive` profile also appears to be completely distinct between these.

## Conclusions

Given the differences between primary/metastasis references for both patients and xenografts, it seems worth doing some initial exploration of annotation performance with `SingleR` next.

## Session Info

```{r session info}
sessionInfo()
```
