---
params: 
  dataset: "hm-6k"
title: "`r glue::glue('Doublet benchmarking results: {params$dataset}')`"
author: Stephanie J. Spielman
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: true
    toc_depth: 4
    code_folding: hide
---

The goal of this notebook is to explore doublet inferences run on ground-truth benchmarking data. 
Importantly, the ground-truth doublets in these benchmarking datasets were themselves called with computational approaches in the context of multiplexed experiments.
This means the true values are not necessarily entirely accurate.

These three doublet detection methods are explored:

- `scDblFinder`
- `scrublet`
- A version of `cxds` [calculated by `scDblFinder`](https://github.com/plger/scDblFinder/blob/893c958322df5a4b3d364f28ed818c740de16316/R/misc.R#L268-L274).
Unlike the other two methods, this method reports only a score and does not attempt to make single/doublet calls, so we use thresholds of 0.5, 0.75, and 0.9 threshold to make `cxds` calls.

This notebook explores inferences made on a single dataset at a time (here, **`r params$dataset`**) with the goal of examining the relationship between scores and accuracy of calls (if made) compared to the ground-truth values.
Note that ground-truth calls are available only for heterotypic doublets, which these methods are more tuned for in the first place (vs homotypic).

## Dataset description

The following dataset description is adapted from [Xi and Li (2021)](https://doi.org/10.1016/j.cels.2020.11.008).

```{r results = "asis"}
if (params$dataset == "hm-6k") {
  data_desc <- "The `hm-6k` dataset is a mixture of human HEK293T and mouse NIH3T3 cells with 6806 droplets."
} else if (params$dataset == "HMEC-orig-MULTI") {
  data_desc <- "The `HMEC-orig-MULTI` dataset contains human primary mammary epithelial cells (HMECs) with 26426 droplets."
} else if (params$dataset == "pdx-MULTI") {
  data_desc <- "The `pdx-MULTI` dataset is a mixture of human breast cancer cells and mouse immune cells from a PDX mouse model."
} else if (params$dataset == "pbmc-1B-dm") {
  data_desc <- "The `pbmc-1B-dm` dataset is a PMBC sample from a patient with systemic lupus erythematosus."
} else {
  stop("Unknown dataset provided.")
}

data_desc |>
  glue::glue() |>
  print()
```



## Setup

### Packages


```{r packages}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(readr)
  library(ggplot2)
  library(patchwork)
  library(caret)
})

theme_set(theme_bw())
```

### Paths


#### Base directories

```{r base paths}
module_base <- rprojroot::find_root(rprojroot::is_renv_project)
data_dir <- file.path(module_base, "scratch", "benchmark-datasets", params$dataset)
result_dir <- file.path(module_base, "results", "benchmark-results")
```

#### Input and output files

Set paths to input and output directories and files in the chunk below.

```{r paths}
# doublet calls
scdbl_tsv <- file.path(
  result_dir, 
  glue::glue("{params$dataset}_scdblfinder.tsv")
)
scrub_tsv <- file.path(
  result_dir, 
  glue::glue("{params$dataset}_scrublet.tsv")
)

# SCE, with ground truth included
sce_file <- file.path(
  data_dir, 
  glue::glue("{params$dataset}.rds")
)

```

### Functions

This chunk defines functions to explore doublet detection results.


```{r functions}
print_doublet_table <- function(df, pred_col, is_ground_truth = FALSE) {
  # Function to print a table of doublet counts and a sentence stating the percentage of doublets
  # df is expected to contain `pred_col`, should should be provided as a string
  
  # If there are no predicted doublets, make sure the tab has a 0 count
  dub_tab <- table( factor(df[[pred_col]], levels = c("singlet", "doublet") ))
  dub_percent <- round(dub_tab[["doublet"]]/sum(dub_tab), 4) * 100
  
  if (is_ground_truth) {
    output <- glue::glue("{dub_percent}% of droplets in this dataset are doublets.") 
  } else {
    output <- glue::glue("{dub_percent}% of droplets in this dataset were predicted to be doublets.")
  }
  
  return(
    list(
      "table" = dub_tab, 
      "percent" = output
    )
  )
}

plot_pca <- function(df, 
                     color_column, 
                     pred_column = NULL,
                     color_type = "d",
                     color_lab) {
  # Plot PCs colored by either discrete or continuous color variable, showing doublets on top
  # df is expected to contain columns PC1, PC2, `color_column`, and if color_type == "c" also `pred_column`. These should _not_ be provided as strings
  # By default, this function assumes a discrete color variable (`color_type = "d"`) with two levels. Use `color_type = "c"` for continuous. 
  
  p <- ggplot(df) + 
    aes(x = PC1, 
        y = PC2, 
        color = {{color_column}}) +
  geom_point(alpha = 0.5) + 
  labs(color = color_lab) + 
  theme(
    legend.title.position = "top",
    legend.position = "bottom"
  )
  
  # Set the palette, and ensure doublets are on top for visibility
  if (color_type == "d") {
    p <- p + 
      scale_color_manual(values = c("black", "lightblue")) + 
    geom_point(
      data = dplyr::filter(df, {{color_column}} == "doublet"), 
      color = "black"
    )
  } else if (color_type == "c") {
    p <- p + 
      scale_color_viridis_c(direction = -1) + 
    geom_point(
      data = dplyr::filter(df, {{pred_column}} == "doublet"), 
      aes(color = {{color_column}})
    )
  }
  
  return(p)
}



plot_jitter_scores <- function(df, score_col, pred_col, method_name){
  # Plot jitter plot of score distributions and predictions, colored by ground truth
  # df is expected to contain columns `ground_truth`, `score_col`, and `pred_col`
  # The column arguments should _not_ be provided as strings
  ggplot(df) +
    aes(x = ground_truth,
        y = {{score_col}}, 
        color = {{pred_col}}) + 
    geom_jitter(
      alpha = 0.2, 
      width = 0.1
    ) + 
    labs(
      x = "Ground truth",
      y = glue::glue("{method_name} score"),
      color = glue::glue("{method_name} prediction")
    ) + 
  theme(
    legend.title.position = "top",
    legend.position = "bottom"
  )
}


plot_density_scores <- function(df, score_col, method_name){
  # Plot density plot of score distributions, colored by ground truth
  # df is expected to contain columns `ground_truth` and `score_col`
  # The column arguments should _not_ be provided as strings
  ggplot(df) +
    aes(x = {{score_col}}, 
        fill = ground_truth) + 
    geom_density(alpha = 0.7) + 
    labs(
      x = glue::glue("{method_name} score"),
      fill = "Ground truth call"
    ) + 
  theme(
    legend.title.position = "top",
    legend.position = "bottom"
  )
}

assess_doublets <- function(df, pred_col) {
  # df is expected to contain columns `ground_truth` and `pred_col`
  # The `pred_col` argument _should_ be provided as a strings
  # This function will return NULL if there are insufficient counts to calculate metrics
  
    # truth should be first
    confusion_table <- table(
      "Truth" = df$ground_truth,
      "Prediction" = df[[pred_col]]
    )
    
    if (!all(dim(confusion_table) == c(2,2))) {
      print("Metrics could not be calculated; insufficient counts.")
      return(NULL)
    } else {
      return(
        caret::confusionMatrix(
          confusion_table, 
          positive = "doublet"
        )
      )
    }
    
}

```


### Read and prepare input data

First, we'll read in and combine TSV files with doublet results.

```{r}
scdbl_df <- read_tsv(scdbl_tsv) |>
  dplyr::select(
    barcodes,
    cxds_score, 
    scdbl_score = score, 
    scdbl_prediction  = class
  ) |>
  # add cxds calls at three thresholds:
  # 0.5, 0.75, 0.9
  dplyr::mutate(
    cxds_prediction_0.5 = dplyr::if_else(
      cxds_score >= 0.5,
      "doublet",
      "singlet"
    ),
    cxds_prediction_0.75 = dplyr::if_else(
      cxds_score >= 0.75,
      "doublet",
      "singlet"
    ),
    cxds_prediction_0.9 = dplyr::if_else(
      cxds_score >= 0.9,
      "doublet",
      "singlet"
    )
  )

scrub_df <- read_tsv(scrub_tsv) 

# grab ground truth and PCA coordinates
sce <- read_rds(sce_file)
sce_df <- scuttle::makePerCellDF(sce, use.dimred = "PCA") |>
  tibble::rownames_to_column(var = "barcodes") |>
  dplyr::select(barcodes,
                ground_truth = ground_truth_doublets, 
                PC1 = PCA.1, 
                PC2 = PCA.2)

# we can now remove the sce
rm(sce)

doublet_df <- dplyr::left_join(
  scdbl_df,
  scrub_df, 
  by = "barcodes"
) |>
  dplyr::left_join(
    sce_df, 
    by = "barcodes"
  )
```


## Ground truth

First, we assess the ground truth doublet distributions to contextualize the doublet calls.


```{r}
dub_tab <- print_doublet_table(doublet_df, "ground_truth", is_ground_truth = TRUE)
dub_tab$table
```

```{r results = "asis"}
dub_tab$percent
```

                     
```{r}
plot_pca(doublet_df, ground_truth, color_lab = "Ground truth doublets")
```


## Doublet predictions and score distributions

Both `scDblFinder` and `scrublet` make doublet calls based on an internally-determined threshold, which may not be optimal. 
These sections explore the score distributions and how they relate to calls, including:

- PCAs of predicted doublet vs. singlets
  - The first PCA is colored by doublet scores, where scores closer to 1 are more likely to be doublets
  - The second PCA is colored by doublet predictions made by the given method and/or threshold.
- Jitter plots of doublet scores
- A confusion with associated performance metrics
  - We're particularly interested in "Balanced accuracy," as there is a much higher proportion of singlets vs doublets.

### scDblFinder results

```{r}
dub_tab <- print_doublet_table(doublet_df, "scdbl_prediction")
dub_tab$table
```

```{r results = "asis"}
dub_tab$percent
```

#### Score distributions relative to ground truth

```{r}
plot_density_scores(
  doublet_df, 
  scdbl_score, 
  "scDblFinder"
)
```

#### PCA plots

```{r fig.width = 8}

d_plot <- plot_pca(
  doublet_df, 
  color_column = scdbl_prediction,
  color_lab = "scDblFinder prediction"
)
  
c_plot <- plot_pca(
  doublet_df, 
  color_column = scdbl_score,
  pred_column = scdbl_prediction,
  color_type = "c",
  color_lab = "scDblFinder score"
)
  
d_plot + c_plot
```

#### Jitter plot

```{r}
plot_jitter_scores(
  doublet_df,
  scdbl_score,
  scdbl_prediction,
  "scDblFinder prediction"
)
```


#### Confusion matrix and performance metrics

```{r}
assess_doublets(
  doublet_df,
  "scdbl_prediction"
) |>
  print()
```


### scrublet results


```{r}
dub_tab <- print_doublet_table(doublet_df, "scrublet_prediction")
dub_tab$table
```

```{r results = "asis"}
dub_tab$percent
```

#### Score distributions relative to ground truth

```{r}
plot_density_scores(
  doublet_df, 
  scrublet_score, 
  "scrublet"
)
```


#### PCA plots

```{r fig.width = 8}
d_plot <- plot_pca(
  doublet_df, 
  color_column = scrublet_prediction,
  color_lab = "scrublet prediction"
)
  
c_plot <- plot_pca(
  doublet_df, 
  color_column = scrublet_score,
  pred_column = scrublet_prediction,
  color_type = "c",
  color_lab = "scrublet score"
)
  
d_plot + c_plot
```

#### Jitter plot

```{r}
plot_jitter_scores(
  doublet_df,
  scrublet_score,
  scrublet_prediction,
  "scrublet prediction"
)
```


#### Confusion matrix and performance metrics

```{r}
assess_doublets(
  doublet_df,
  "scrublet_prediction"
) |>
  print()
```


### cxds results


Since this method does not call droplets as singlet/doublet, performance is assessed below at several thresholds: 0.5, 0.75, and 0.9.


#### Score distributions relative to ground truth

```{r}
plot_density_scores(
  doublet_df, 
  cxds_score, 
  "cxds"
)
```

#### PCA plots

```{r fig.width = 10}

cxds_cols <- names(doublet_df)[grep("^cxds_prediction_", names(doublet_df))]

prediction_pcas <- cxds_cols |>
  purrr::map(
    \(col_name) {
      threshold <- stringr::str_remove(col_name, "cxds_prediction_")
        
      plot_pca(
        doublet_df, 
        color_column = !!as.symbol(col_name),
        color_lab = glue::glue("cxds prediction at {threshold}")
      ) +
      labs(title = glue::glue("cxds prediction at {threshold}"))
    }
)
wrap_plots(prediction_pcas)
```

```{r}
# just need a single score pca for this case
# we'll provide the least stringent predictions; this only affects point layering in the plot
plot_pca(
  doublet_df, 
  color_column = cxds_score, 
  pred_column = cxds_prediction_0.5,
  color_type = "c",
  color_lab = glue::glue("cxds scores")
)
```



#### cxds: 0.5 threshold

```{r}
dub_tab <- print_doublet_table(doublet_df, "cxds_prediction_0.5")
dub_tab$table
```

```{r results = "asis"}
dub_tab$percent
```

##### Jitter plot

```{r}
plot_jitter_scores(
  doublet_df,
  cxds_score,
  cxds_prediction_0.5,
  "cxds prediction at 0.5"
)
```


##### Confusion matrix and performance metrics

```{r}
assess_doublets(
  doublet_df,
  "cxds_prediction_0.5"
) |>
  print()
```


#### cxds: 0.75 threshold

```{r}
dub_tab <- print_doublet_table(doublet_df, "cxds_prediction_0.75")
dub_tab$table
```

```{r results = "asis"}
dub_tab$percent
```

##### Jitter plot


```{r}
plot_jitter_scores(
  doublet_df,
  cxds_score,
  cxds_prediction_0.75,
  "cxds prediction at 0.75"
)
```


##### Confusion matrix and performance metrics

```{r}
assess_doublets(
  doublet_df,
  "cxds_prediction_0.75"
) |>
  print()
```

#### cxds: 0.9 threshold

```{r}
dub_tab <- print_doublet_table(doublet_df, "cxds_prediction_0.9")
dub_tab$table
```

```{r results = "asis"}
dub_tab$percent
```

##### Jitter plot

```{r}
plot_jitter_scores(
  doublet_df,
  cxds_score,
  cxds_prediction_0.9,
  "cxds prediction at 0.9"
)
```

##### Confusion matrix and performance metrics

```{r}
assess_doublets(
  doublet_df,
  "cxds_prediction_0.9"
) |>
  print()
```


## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```
