---
params: 
  dataset: "hm-6k"
title: "`r glue::glue('Doublet benchmarking results: {params$dataset}')`"
author: Stephanie J. Spielman
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: true
---

The goal of this notebook is to explore doublet inferences run on ground-truth benchmarking data. 
These three doublet detection methods are explored:

- `scDblFinder`
- `scrublet`
- A version of `cxds` calculated by `scDblFinder`.
Unlike the other two methods, this one reports only a score and does not attempt to make single/doublet calls, so we use thresholds of 0.5, 0.75, and 0.9 threshold to make `cxds` calls.

This notebook explores inferences made on a single dataset at a time (here, **`r params$dataset`**) with the goal of examining the relationship between scores and accuracy of calls (if made) compared to the ground-truth values.
Note that ground-truth calls are available only for heterotypic doublets, which these methods are more tuned for in the first place (vs homotypic).


## Setup

### Packages


```{r packages}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(readr)
  library(ggplot2)
  library(caret)
})

theme_set(theme_bw())
```

### Paths


#### Base directories

```{r base paths}
module_base <- rprojroot::find_root(rprojroot::is_renv_project)
data_dir <- file.path(module_base, "scratch", "benchmark-datasets", params$dataset)
result_dir <- file.path(module_base, "results", "benchmark-results")
```

#### Input and output files

Set paths to input and output directories and files in the chunk below.

```{r paths}
# Input files
scdbl_tsv <- file.path(
  result_dir, 
  glue::glue("{params$dataset}_scdblfinder.tsv")
)
scrub_tsv <- file.path(
  result_dir, 
  glue::glue("{params$dataset}_scrublet.tsv")
)
truth_tsv <- file.path(
  data_dir, 
  glue::glue("{params$dataset}_ground-truth.tsv")
)
```

### Functions

This chunk defines functions to explore doublet detection results


```{r functions}
plot_doublets <- function(df, score_col, pred_col, method_name){
  # df is expected to contain columns `ground_truth`, `score_col`, and `pred_col`
  # The column arguments should _not_ be provided as strings
  ggplot(df) +
    aes(x = ground_truth,
        y = {{score_col}}, 
        color = {{pred_col}}) + 
    geom_jitter(
      alpha = 0.2, 
      width = 0.1
    ) + 
    labs(
      x = "Ground truth call",
      y = glue::glue("{method_name} score"),
      color = glue::glue("{method_name} prediction")
    )
}


assess_doublets <- function(df, pred_col) {
  # df is expected to contain columns `ground_truth` and `pred_col`
  # The `pred_col` argument _should_ be provided as a strings
  
  caret::confusionMatrix(
    # truth should be first
    table(
      "Truth" = doublet_df$ground_truth,
      "Prediction" = doublet_df[[pred_col]]
    ), 
    positive = "doublet"
  )
  
}

```


### Read and prepare input data

First, we'll read in and combine TSV files with doublet results.

```{r}
scdbl_df <- read_tsv(scdbl_tsv) |>
  dplyr::select(
    barcodes,
    cxds_score, 
    scdbl_score = score, 
    scdbl_prediction  = class
  ) |>
  # add cxds calls at three thresholds:
  # 0.5, 0.75, 0.9
  dplyr::mutate(
    cxds_prediction_0.5 = dplyr::if_else(
      cxds_score >= 0.5,
      "doublet",
      "singlet"
    ),
    cxds_prediction_0.75 = dplyr::if_else(
      cxds_score >= 0.75,
      "doublet",
      "singlet"
    ),
    cxds_prediction_0.9 = dplyr::if_else(
      cxds_score >= 0.9,
      "doublet",
      "singlet"
    )
  )

scrub_df <- read_tsv(scrub_tsv) 

true_df <- read_tsv(truth_tsv)

doublet_df <- dplyr::inner_join(
  scdbl_df,
  scrub_df, 
  by = "barcodes"
) |>
  dplyr::inner_join(
    true_df, 
    by = "barcodes"
  )
```

## Explore results: Score distributions

Both `scDblFinder` and `scrublet` make doublet calls based on an internally-determined threshold, which may not be optimal. 

These sections explore the score distributions and how they relate to calls, including a confusion matrix and associated performance metrics. 
We're particularly interested in "Balanced accuracy," as there is a much higher proportion of singlets vs doublets.

### scDblFinder results

```{r}
plot_doublets(
  doublet_df,
  scdbl_score,
  scdbl_prediction,
  "scDblFinder"
)

assess_doublets(
  doublet_df,
  "scdbl_prediction"
) |>
  print()
```


### scrublet results

```{r}
plot_doublets(
  doublet_df,
  scrublet_score,
  scrublet_prediction,
  "scrublet"
)

assess_doublets(
  doublet_df,
  "scrublet_prediction"
) |>
  print()
```


### cxds results

> This method does not call droplets as singlet/doublet.
> Instead, performance is assessed below at several thresholds:

#### cxds: 0.5 threshold

```{r}
plot_doublets(
  doublet_df,
  cxds_score,
  cxds_prediction_0.5,
  "cxds, 0.5"
)

assess_doublets(
  doublet_df,
  "cxds_prediction_0.5"
) |>
  print()
```


#### cxds: 0.75 threshold

```{r}
plot_doublets(
  doublet_df,
  cxds_score,
  cxds_prediction_0.75,
  "cxds, 0.75"
)

assess_doublets(
  doublet_df,
  "cxds_prediction_0.75"
) |>
  print()
```

#### cxds: 0.9 threshold

```{r}
plot_doublets(
  doublet_df,
  cxds_score,
  cxds_prediction_0.9,
  "cxds, 0.9"
)

assess_doublets(
  doublet_df,
  "cxds_prediction_0.9"
) |>
  print()
```


## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```
