---
title: "Identify max rank for AUCell"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
---

This notebook is used to find the appropriate `aucMaxRank` to use with `AUCell` for all samples in `SCPCP00015`. 

From the  [`AUCell` vignette](https://bioconductor.org/packages/release/bioc/vignettes/AUCell/inst/doc/AUCell.html):

> It is important to check that most cells have at least the number of expressed/detected genes that are going to be used to calculate the AUC (aucMaxRank in calcAUC()). The histogram provided by AUCell_buildRankings() allows to quickly check this distribution. 

Here we will build the rankings and look at the histogram showing the number of genes detected per cell for each sample and use that information to pick a threshold (percentage of genes) to use across all samples. 

## Setup

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_classic()
)

# set seed
set.seed(2025)
```


```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current", "SCPCP000015")

# paths to all processed sce files
sce_files <- list.files(data_dir, pattern = "*processed.rds", full.names = TRUE, recursive = TRUE)
```

```{r}
# read in SCE files 
sce_list <- sce_files |> 
  purrr::map(readr::read_rds)
```


## Filter `SingleCellExperiment` objects

The default `aucMaxRank` is calculated as `ceiling(0.05*nrow(exprMat))` where the `exprMat` is the counts matrix. 
For the processed `SingleCellExperiment` objects in ScPCA we include all genes in the index, regardless of if they are detected in our sample. 
This means the rankings might include a bunch of genes that aren't actually present and the `aucMaxRank` would be calculated based on a set of genes that aren't even there, setting the threshold higher than it should be. 
Because of this, we probably want to filter the `SingleCellExperiment` objects to remove any genes that are not detected in any cells prior to calculating ranks or running `AUCell`. 

```{r}
# get a list of filtered sce
# remove any genes that are not detected in any cells 
sce_list <- sce_list |> 
  purrr::map(\(sce){
    
    genes_to_remove <- rowData(sce)$detected > 0 
    filtered_sce <- sce[rowData(sce)$detected > 0 , ]
    
  })
```

## `AUCell` build rankings 

Below we will look at the distribution of the number of genes detected per cell for each sample using the `AUCell_buildRankings()` function. 


```{r}
cell_rankings_list <- sce_list |> 
  purrr::map(\(sce){AUCell::AUCell_buildRankings(counts(sce))})
```

Now we will look at the potential thresholds for `aucMaxRank` using 5% and 1% of genes. 
The numbers printed here represent the total number of genes that would be used as the `aucMaxRank` for a given dataset using the threshold. 
We can then use the histograms to see how many cells have at least that number of genes detected. 
Remember, we want to pick a number of genes that are detected in most cells.
If we picked a max rank higher than the number of genes detected in most cells, the non-detected genes that are randomly ordered would play an outsized role in our AUC values.

### 5%  

```{r}
sce_list |> 
  purrr::map_dbl(\(sce){ceiling(nrow(counts(sce))*0.05)})
```

This looks too high based on our histograms. 
We would have a lot of cells where the number of genes detected is lower than the max rank, which is not ideal. 

### 1% 

```{r}
sce_list |> 
  purrr::map_dbl(\(sce){ceiling(nrow(counts(sce))*0.01)})
```

This looks better to me. 
I also think there's enough variation in some of the samples that we don't want to just pick one max rank for all samples, but instead use a percentage of the total genes detected. 

## Conclusions

- Prior to running `AUCell` we should remove any genes that are not detected in any cells.
- Using a cutoff of 1% of the total genes detected seems like a reasonable choice for the `aucMaxRank`. 

## Session info 

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```


