---
title: "Explore CNV distributions of reference cells: SCPCP000015"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
---


## Introduction

We have previously observed that the reference cell CNV distributions in pooled references may overlap with tumor CNV distributions.
This notebook aims to explore the extent of this trend for `SCPCP000015`.

## Setup

```{r}
options(readr.show_col_types = FALSE)
suppressPackageStartupMessages({
  library(ggplot2)
  library(SingleCellExperiment)
})

theme_set(theme_bw())
```

### Functions

This function prepares a table of `inferCNV` results by creating a `total_cnv_per_cell` column to summarize CNVs per cell. 

```{r}
prepare_cnv_df <- function(infercnv_tsv) {
  infercnv_df <- readr::read_tsv(infercnv_tsv)

  infercnv_metadata_df <- infercnv_df |>
    dplyr::select(
      -starts_with("has_"),
      -starts_with("proportion_"),
      -starts_with("top_")
    ) |>
    dplyr::mutate(cell_group = stringr::str_split_i(subcluster, "_", 1))

  infercnv_df |>
    tidyr::pivot_longer(
      starts_with("has_cnv_"),
      names_to = "chr",
      values_to = "cnv"
    ) |>
    # sum all values for each
    dplyr::group_by(cell_id) |>
    dplyr::summarize(total_cnv_per_cell = sum(cnv)) |>
    dplyr::ungroup() |>
    # bring back metadata
    dplyr::inner_join(
      infercnv_metadata_df,
      by = "cell_id"
    )
}
```


This function plots ridge plots of total CNV values across cell groupings for a given normal reference.

```{r}
plot_cnv_distributions <- function(df, reference_name) {
  filtered_df <- infercnv_df |>
    dplyr::filter(normal_reference == reference_name)

  ggplot(filtered_df) +
    aes(x = total_cnv_per_cell, y = label, fill = label) +
    ggridges::geom_density_ridges2(scale = 0.9) +
    facet_wrap(vars(infercnv_library_id)) +
    theme(legend.position = "none") +
    ggtitle(reference_name)
}
```


### Paths

#### Base directories


```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

module_dir <- file.path(repository_base, "analyses", "infercnv-consensus-cell-type")
result_dir <- file.path(module_dir, "results", "SCPCP000015") # organized as sample/reference/infercnv results
ewings_dir <- file.path(repository_base, "data", "current", "results", "cell-type-ewings", "SCPCP000015")
```

#### Input files

```{r}
# inferCNV output files
infercnv_files <- list.files(
  path = result_dir,
  pattern = "_cnv-metadata\\.tsv",
  recursive = TRUE,
  full.names = TRUE
) |>
  # remove any containing "internal"; we only want pooled
  purrr::discard(\(x) {
    stringr::str_detect(x, "internal")
  }) |>
  # name by "library-reference"
  purrr::set_names(\(x) {
    library_id <- stringr::str_split_i(basename(x), "_", 1)
  })


# ewing module cell types
celltype_files <- list.files(
  ewings_dir,
  pattern = "_ewing-celltype-assignments\\.tsv",
  recursive = TRUE,
  full.names = TRUE
) |>
  # name by library id
  purrr::set_names(\(x) {
    stringr::str_split_i(basename(x), "_", 1)
  })
```


### Read input files


We'll first read in the cell type annotations:
```{r}
celltype_df <- celltype_files |>
  purrr::map(readr::read_tsv) |>
  purrr::list_rbind(names_to = "library_id") |>
  dplyr::mutate(cell_id = glue::glue("{library_id}-{barcodes}")) |>
  dplyr::select(cell_id, consensus_annotation, ewing_annotation)
```


Now, we'll read in the `inferCNV` results:
```{r}
infercnv_df <- infercnv_files |>
  purrr::map(prepare_cnv_df) |>
  purrr::list_rbind(names_to = "infercnv_library_id") |>
  dplyr::inner_join(celltype_df, by = "cell_id") |>
  # condense labels for cell type groupings
  dplyr::mutate(
    label = ifelse(
      stringr::str_detect(ewing_annotation, "tumor"),
      "tumor",
      cell_group
    ),
    label = forcats::fct_relevel(label, "unknown", "tumor", "reference")
  )
```


## Analysis

We'll first visualize the total CNV distributions for each inference for each grouping of cells:

- `unknown`: query cells
- `tumor`: query cells labeled "tumor" by `cell-type-ewings`
- `reference`: cells in the normal reference

```{r, fig.height = 8}
plot_cnv_distributions(infercnv_df, "endo_pooled")
plot_cnv_distributions(infercnv_df, "immune_pooled")
plot_cnv_distributions(infercnv_df, "endo-immune_pooled")
```


* The `endo` reference cell CNV distributions generally do not overlap with tumor cell CNV distributions.
* The `endo-immune` and `immune` reference cell CNV distributions do have a wider spread but do not _substantially_ overlap with the tumor cell CNV distributions. 

Let's look more closely at the `immune` reference to see if any particular libraries might be driving a (minor) trend.
For each inference with the pooled `immune` reference, we'll find the 25\% top reference cells with the highest CNVs.
Do these cells tend to be from a particular library?

First, note how many reference cells come from each library:

```{r}
ref_immune_df <- infercnv_df |>
  dplyr::filter(
    normal_reference == "immune_pooled",
    cell_group == "reference"
  ) |>
  tidyr::separate(cell_id, into = c("reference_library_id", "barcode"), sep = "-", remove = FALSE)


ref_immune_df |>
  dplyr::select(cell_id) |>
  dplyr::distinct() |>
  tidyr::separate(cell_id, into = c("reference_library_id", "barcode"), sep = "-") |>
  dplyr::count(reference_library_id)
```

The table displayed below shows for each inference (`infercnv_library_id`), how many of top 25\% CNV cells in the pooled immune reference are from each library.
For example, of the reference cells with the highest CNVs in the `SCPCL000822` inference, 28 cells came from `SCPCL000822`, 30 cells came from `SCPCL000824`, etc.
Generally, these numbers are consistent across inferences although cells from `SCPCL000824` do seem to be more prevalent in certain samples.
Even so, there doesn't seem to be any major trend here.

```{r}
top25 <- ref_immune_df |>
  dplyr::group_by(infercnv_library_id) |>
  dplyr::slice_max(total_cnv_per_cell, prop = 0.25) |>
  dplyr::ungroup()
```



```{r}
top25 |>
  dplyr::count(infercnv_library_id, reference_library_id) |>
  tidyr::pivot_wider(
    names_from = reference_library_id,
    values_from = n,
    values_fill = 0
  )
```

## Conclusions

It does not seem that reference cell CNV distributions are problematically overlapping with tumor CNV distributions now that references were updated to generate these re-processed results.

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```
