---
title: "Comparison of clusters to cell types for `r params$library_id`"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
params:
  library_id: "SCPCL000822"
  sce_file: "../../../../data/current/SCPCP000015/SCPCS000490/SCPCL000822_processed.rds"
  singler_results_file: "../../results/aucell_singler_annotation/SCPCS000490/SCPCL000822_singler-classifications.tsv"
  cluster_results_file: "../../results/cluster_exploration/SCPCS000490/SCPCL000822_clusters.tsv"
  marker_genes_file: "../../references/visser-all-marker-genes.tsv"
---

TODO: Update this notebook to take as input a clustering algorithm, nn (range of nn?), resolution, and optional objective function to display in SingleR and marker gene plots

- The clusters are then compared to the results from running `SingleR` in the `aucell-singler-annotation.sh` workflow.
- We then look at marker gene expression across the clusters and assign each cluster to a cell type. 

## Setup

```{r}
# check that sce and results files exist
stopifnot(
  "sce file does not exist" = file.exists(params$sce_file),
  "SingleR results file does not exist" = file.exists(params$singler_results_file),
  "Marker genes file does not exist" = file.exists(params$marker_genes_file)
)
```


```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_classic()
)

# quiet messages
options(readr.show_col_types = FALSE)
ComplexHeatmap::ht_opt(message = FALSE)

# set seed
set.seed(2024)
```


```{r base paths}
# The path to this module
module_base <- rprojroot::find_root(rprojroot::is_renv_project)

# source in helper functions: calculate_sum_markers()
validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
source(validation_functions)

# source in clustering functions
clustering_functions <- file.path(module_base, "scripts", "utils", "clustering-functions.R")
source(clustering_functions)
```


```{r}
# read in input sce files
sce <- readr::read_rds(params$sce_file)

# read in singler results
singler_classification_df <- readr::read_tsv(params$singler_results_file) |>
  dplyr::rename("cell_id" = "barcodes")
```


## Clusters vs. `SingleR` {.tabset}

Now we will compare the clustering results to the cell type assignments obtained from `SingleR` in the `aucell-singler-annotation.sh` workflow. 
To compare results we will calculate the Jaccard similarity index between clusters and cell types. 
We expect that good clustering will line up with the cell type assignments so that there is ~ 1 cell type per cluster. 

For the plots, we will only display the top 7 cell types identified by `SingleR` and all other cells will be lumped together into `All remaining cell types`. 

```{r}
cluster_classification_df <- all_cluster_results |>
  # add in classifications from singler
  dplyr::left_join(singler_classification_df, by = "cell_id") |>
  dplyr::mutate(
    # first grab anything that is tumor and label it tumor
    # NA should be unknown
    singler_annotation = dplyr::case_when(
      stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
      is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
      .default = singler_annotation
    ) |>
      forcats::fct_relevel("tumor", after = 0),
    # get the top cell types for plotting later
    singler_lumped = singler_annotation |>
      forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
      forcats::fct_infreq() |>
      forcats::fct_relevel("All remaining cell types", after = Inf)
  )

cluster_classification_df |> 
  split(cluster_classification_df$resolution) |> 
  purrr::map(\(df){cluster_celltype_heatmap(df)}) 

```

```{r, fig.height=10, fig.width=7}
# get heatmap for showing cluster assignments vs. singler assignments
cluster_celltype_heatmap(cluster_classification_df)
```


## Marker gene expression {.tabset} 

Finally, we will look at the expression of marker genes across each cluster for each value of `k`. 
In these plots, each row shows the distribution of the specified marker genes in that cluster. 
Each panel is labeled with the marker gene expression being plotted. 

```{r}
# read in marker genes table
marker_genes_df <- readr::read_tsv(params$marker_genes_file) |>
  # account for genes being from multiple sources
  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |>
  dplyr::distinct()

# get list of all cell types found
cell_types <- unique(marker_genes_df$cell_type)

# get the sum of expression of all genes for each cell type
gene_exp_df <- cell_types |>
  purrr::map(\(type){
    calculate_sum_markers(marker_genes_df, sce, type)
  }) |>
  purrr::reduce(dplyr::inner_join, by = "barcodes")

# join sum expression columns with clustering results
cluster_exp_df <- all_cluster_results |>
  dplyr::left_join(gene_exp_df, by = "barcodes")
```

```{r, fig.height=7, fig.width=7}
# plot to look at marker gene expression for all cell types across all clusters and all params
k_params <- unique(cluster_exp_df$nn_param) |>
  as.character()

k_params |>
  purrr::map(\(k) {
    plot_marker_genes(cluster_exp_df,
      k_value = k
    )
  })
```


## Session info 

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```
