---
title: "Exploration of clustering results"
author: maud-p
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: kable
params:
  sample_name: ~
  sample_path: ~
  out_rds_path: ~
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warnings = FALSE,
	error = TRUE
)
```


## Introduction

This a template for an analysis notebook using RMarkdown.

In this notebook, we will set up parameters in the seurat workflow [normalization --> reduction --> clustering] for one Wilms tumor sample (SCPCS000169) of the Wilms Tumor dataset (SCPCP000006). 

This correspond to the step 2 of the proposed analysis: clustering of cells across a set of parameters for few samples

- https://github.com/AlexsLemonade/OpenScPCA-analysis/discussions/635
- https://github.com/AlexsLemonade/OpenScPCA-analysis/issues/671

In order to assess the clustering quality, we look into some marker genes, pathways enrichment and label transfer.



## Sample metadata

```{r paths2}
## to be modified when rendered
filedir <- filelist[i] 
sample <- gsub(paste0(data_dir, "/"), "", filedir)

metadata <- read.table(file.path(data_dir, "/single_cell_metadata.tsv"), sep = "\t", header=TRUE)

metadata[metadata$scpca_sample_id == sample,]
```


### Description of the analysis workflow

We perform the following analysis to assess for the quality of clustering:

[1] We perform some quality check to assess any QC-induced clustering (nFeature, nCount, percent.mito). 

[2] We add cell cycle information, as we know that in a specific cell cycle state, the transcriptional program is mostly/exclusively related to cell cycle genes and the identity of cells is difficult to determine. 
We expect these cells to cluster together in a cluster of proliferating cells. 

[3] We look at specific marker genes that we reported in the table marker.sets/CellType_metadata.csv to check the relevance of the clustering.

[4] We look at specific pathways that we reported in the table marker.sets/Pathways_metadata.csv to check the relevance of the clustering.

[5] We run DElegate::FindAllMarkers2 to find markers of the different clusters and manually check if they do make sense. 
DElegate::FindAllMarkers2 is an improved version of Seurat::FindAllMarkers based on pseudobulk differential expression method. 

[6] We perform enrichment analysis of marker genes for each seurat clusters. 
We defined all the genes from the seurat object as the universe and used the MSigDB gene sets. 

[7] We plot pca/umap reduction grouping with available annotations (singler_, cellassign_). 
We expect at least immune cells to be correctly label and fall into a few set of clusters.

[ ] Next step, aim for a next PR: We will label transfer (Azimuth) to transfer annotation from the fetal kidney atlas human reference. 
We plot pca/umap reduction grouping with latest labels. 
We expect it to be the most representative of the cell types in the sample. 


### Selected parameters

Please note: to keep the notebook as straight as possible, we decided to show the analysis for the selected set of parameters:

- `SCTransform` (default parameters)
- `RunPCA` (default) 
- `RunUMAP` (dims = 1:50) 
- `FindNeighbors` (dims = 1:50) 
- `FindClusters` (default)


Note: Other parameters have been previously tested, but we would like to show in the following report that the one selected is performing good.


## Analysis content

### Load the processed SingleCellExperiment rds object 


```{r load}
file <- list.files(filedir)
file <- file[grepl("_processed.rds", file)]

# open the processed rds object
sce <- readRDS(paste0(filedir, "/", file))
```


### Run Seurat pipeline

```{r seurat_wf, fig.width=8, fig.height=4, out.width='100%'}
# convert to seurat
srat <- CreateSeuratObject(counts = counts(sce),
                                    assay = "RNA",
                                    project = sample
                           )

# convert colData and rowData to data.frame for use in the Seurat object
cell_metadata <- as.data.frame(colData(sce))
row_metadata <- as.data.frame(rowData(sce))

# add cell metadata (colData) from SingleCellExperiment to Seurat
srat@meta.data <- cell_metadata

# add row metadata (rowData) from SingleCellExperiment to Seurat
srat[["RNA"]]@meta.data <- row_metadata

# add metadata from SingleCellExperiment to Seurat
srat@misc <- metadata(sce)
rm(sce)

# Normalization
options(future.globals.maxSize= 8912896000000)
srat <- SCTransform(srat, verbose = F, conserve.memory = TRUE)

# dimensionality reduction
srat <- RunPCA(srat, verbose = F)
srat <- RunUMAP(srat, dims = 1:50, verbose = F)

# clustering
srat    <- FindNeighbors(srat, dims = 1:50, verbose = F)
srat    <- FindClusters(srat, verbose = T)
  
```



### Visualize seurat clusters

```{r fig.width=20, fig.height=5, out.width='100%'}

d2 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - umap")
d1 <- SCpubr::do_DimPlot(srat, reduction="pca", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - pca")
v <- SCpubr::do_ViolinPlot(srat, features = c( "subsets_mito_percent"), ncol = 1, group.by = "seurat_clusters", legend.position = "none") 

d1 + d2 + v + plot_layout(ncol = 3, widths = c(2,2,4))

```

We expect up to 5 set of clusters:

- blastema cancer cells
- epithelium cancer and/or normal cells
- stroma cancer and/or normal cells
- immune cells
- endothelial cells





### Cell cycle information
```{r cell_cycle, fig.height=20, fig.width=30}
s.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$s.genes]
g2m.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$g2m.genes]

srat <- CellCycleScoring(srat, s.features = s.genes, g2m.features = g2m.genes, set.ident = FALSE)
```

```{r fig.width=20, fig.height=5, out.width='100%'}

d2 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "Phase", label = TRUE) + ggtitle("Phase - umap")
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - umap")

b <- SCpubr::do_BarPlot(sample = srat,
                         group.by = "Phase",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file)
d1 + d2 + b + plot_layout(ncol = 3, widths = c(2,2,4))


```

```{r fig.width=10, fig.height=2.5, out.width='100%'}
 SCpubr::do_ViolinPlot(srat, features = c("S.Score", "G2M.Score"), ncol = 2, group.by = "seurat_clusters", legend.position = "none") 
```




### Look at specific genes

Here, we open the table of marker genes marker-sets/CellType_metadata.csv. 
Note: we do not expect to have a clear and nice pattern of expression for all of the following markers in every tumor. 
This is just ti get a few idea.

```{r marker_genes, fig.width=12, fig.height=6, out.width='100%'}

DT::datatable(CellType_metadata, caption = ("CellType_metadata"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

                               buttons = c( 'csv', 'excel')))
```

#### Malignant markers


```{r fig.width=25, fig.height=5, out.width='100%'}
 SCpubr::do_ViolinPlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class %in% c("malignant")])] , ncol = 9, group.by = "seurat_clusters", legend.position = "none") 
```

##### NCAM1

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000149294", pt.size = 0.2) + ggtitle("NCAM1")
v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000149294", ncol = 1, group.by = "seurat_clusters", legend.position = "none") 

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```


##### COL6A3

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000163359", pt.size = 0.2) + ggtitle("COL6A3")
v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000163359", ncol = 1, group.by = "seurat_clusters", legend.position = "none") 

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))


```

##### IGF2

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000167244", pt.size = 0.2) + ggtitle("IGF2")
v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000167244", ncol = 1, group.by = "seurat_clusters", legend.position = "none") 

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```


##### WT1

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000184937", pt.size = 0.2) + ggtitle("WT1")
v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000184937", ncol = 1, group.by = "seurat_clusters", legend.position = "none") 

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```

##### EPCAM

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000119888", pt.size = 0.2) + ggtitle("EPCAM")
v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000119888", ncol = 1, group.by = "seurat_clusters", legend.position = "none") 

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```


#### Immune markers

```{r fig.width=20, fig.height=5, out.width='100%'}
 SCpubr::do_ViolinPlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class %in% c("immune")])]
                       , ncol = 6, group.by = "seurat_clusters", legend.position = "none") 
```

##### PTPRC = CD45

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000081237", pt.size = 0.2) + ggtitle("PTPRC = CD45")
v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000081237", ncol = 1, group.by = "seurat_clusters", legend.position = "none", pt.size = 2) 

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```

#### Other markers

```{r fig.width=20, fig.height=5, out.width='100%'}
 SCpubr::do_ViolinPlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[! CellType_metadata$cell_class %in% c("immune", "malignant")])]
                       , ncol = 6, group.by = "seurat_clusters", legend.position = "none") 
```


##### VWF, endothelial cells

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000110799", pt.size = 0.2) + ggtitle("VWF")
v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000110799", ncol = 1, group.by = "seurat_clusters", legend.position = "none", pt.size = 2) 

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```


### Look at specific pathways

Here, we open the table of marker genes marker-sets/Pathway_metadata.csv. 

```{r marker_pathways, fig.width=12, fig.height=6, out.width='100%'}

    
DT::datatable(Pathway_metadata, caption = ("Pathway_metadata"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

                               buttons = c( 'csv', 'excel')))
```

#### TP53 pathway

here we will calculate a TP53 score using AddMduleScore and the genes of the HALLMARK_P53_PATHWAY gene set.

```{r fig.width=10, fig.height=10, out.width='100%'}

## define genesets
hallmarks <- msigdbr(species = "human", category = "H")

TP53_list = hallmarks %>% 
  filter(gs_name == "HALLMARK_P53_PATHWAY") %>%
  dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()

TP53_list <- list(TP53_list$human_ensembl_gene)

srat <- AddModuleScore(srat, features = TP53_list, name = "TP53_score")
```


```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "TP53_score1", pt.size = 0.2) + ggtitle("TP53 score")
v <- SCpubr::do_ViolinPlot(srat, features = "TP53_score1", ncol = 1, group.by = "seurat_clusters", legend.position = "none") 

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```

"Normal" cells (immune, endothelial) have a slightly higher TP53 score. 


#### DNA repair pathway

here we will calculate a DNA_repair score using AddMduleScore and the genes of the HALLMARK_DNA_REPAIR gene set.

```{r fig.width=10, fig.height=10, out.width='100%'}


DNA_repair = hallmarks %>% 
  filter(gs_name == "HALLMARK_DNA_REPAIR") %>%
  dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()

DNA_repair_list <- list(DNA_repair$human_ensembl_gene)

srat <- AddModuleScore(srat, features = DNA_repair_list, name = "DNA_repair_score")
```


```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "DNA_repair_score1", pt.size = 0.2) + ggtitle("DNA_repair score")
v <- SCpubr::do_ViolinPlot(srat, features = "DNA_repair_score1", ncol = 1, group.by = "seurat_clusters", legend.position = "none") 

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```

Note: Chemo-treated samples should have higher DNA-damage scores.


#### DROSHA target genes

```{r fig.width=10, fig.height=10, out.width='100%'}

## define genesets
c3 <- msigdbr(species = "human", category = "C3", subcategory = "TFT:GTRD")

DROSHA_list = c3 %>% 
  filter(gs_name == "DROSHA_TARGET_GENES") %>%
  dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()

DROSHA_list <- list(DROSHA_list$human_ensembl_gene)

srat <- AddModuleScore(srat, features = DROSHA_list, name = "DROSHA_score")
```


```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "DROSHA_score1", pt.size = 0.2) + ggtitle("DROSHA score")
v <- SCpubr::do_ViolinPlot(srat, features = "DROSHA_score1", ncol = 1, group.by = "seurat_clusters", legend.position = "none") 

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```

#### DICER1 target genes

```{r fig.width=10, fig.height=10, out.width='100%'}
DICER1_list = c3 %>% 
  filter(gs_name == "DICER1_TARGET_GENES") %>%
  dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()

DICER1_list <- list(DICER1_list$human_ensembl_gene)

srat <- AddModuleScore(srat, features = DICER1_list, name = "DICER1_score")
```


```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "DICER1_score1", pt.size = 0.2) + ggtitle("DICER1 score")
v <- SCpubr::do_ViolinPlot(srat, features = "DICER1_score1", ncol = 1, group.by = "seurat_clusters", legend.position = "none") 

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```



### Find marker genes for each of the seurat clusters

In addition to the list of known marker genes, we used an unbiased approach to find transcripts that characterized the different clusters.
We run DElegate::FindAllMarkers2 to find markers of the different clusters and manually check if they do make sense. 
DElegate::FindAllMarkers2 is an improved version of Seurat::FindAllMarkers based on pseudobulk differential expression method. 
Please check the preprint from Chistoph Hafemeister: https://www.biorxiv.org/content/10.1101/2023.03.28.534443v1
and tool described here: https://github.com/cancerbits/DElegate 

```{r fig.width=10, fig.height=10, out.width='100%'}
feature_conversion <- srat@assays$RNA@meta.data
de_results   <- DElegate::FindAllMarkers2(srat, group_column = "seurat_clusters")

#filter the most relevant markers
s.markers <- de_results[de_results$padj < cfg$padj_thershold & de_results$log_fc > cfg$lfc_threshold & de_results$rate1 > cfg$rate1_threshold,]

# add gene symbol for easiest interpretation of the result
s.markers$gene_ids <- s.markers$feature
s.markers <- left_join(s.markers,feature_conversion, by = c( "gene_ids") )
identical(s.markers$feature, s.markers$gene_ids)     # check the quality of the merge, must be true

DT::datatable(s.markers, caption = ("marker genes"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

                               buttons = c( 'csv', 'excel')))

# Select top 5 genes for heatmap plotting
s.markers <- na.omit(s.markers)
s.markers %>%
    group_by(group1) %>%
    top_n(n =  5, wt = log_fc) -> top5

# subset for plotting
cells <- WhichCells(srat, downsample = 100)
ss <- subset(srat, cells = cells)
ss <- ScaleData(ss, features = top5$feature)

p1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
p2 <- DoHeatmap(ss, features = top5$feature,  cells = cells, group.by = "seurat_clusters") + NoLegend() + 
  scale_fill_gradientn(colors =  c("#01665e","#35978f",'darkslategray3', "#f7f7f7", "#fee391","#fec44f","#F9AD03")) 
p3 <- ggplot(srat@meta.data, aes(seurat_clusters, fill = seurat_clusters)) + geom_bar() + NoLegend()


common_title <- sprintf("Unsupervised clustering %s, %d cells", srat@meta.data$orig.ident[1], ncol(srat))
show((((p1 / p3) + plot_layout(heights = c(3,2)) | p2) ) + plot_layout(widths = c(1, 2)) + plot_layout(heights = c(3,1)) + plot_annotation(title = common_title))
DT::datatable(top5[, c(1, 9, 11, 12)], caption = ("top 5 marker genes"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

                               buttons = c( 'csv', 'excel')))


```



### Enrichment analysis of marker genes

Here we perform enrichment analysis of the marker genes found in the previous section for each Seurat cluster.

We defined as universe/background all the genes expressed in the dataset, meaning the rownames of the Seurat object.

We used three gene sets from MSiGDB:

- the hallmark gene sets  are coherently expressed signatures derived by aggregating many MSigDB gene sets to represent well-defined biological states or processes.
- the C3 : regulatory target gene sets  based on gene target predictions for microRNA seed sequences and predicted transcription factor binding sites.
- the C8 : cell type signature gene sets  curated from cluster markers identified in single-cell sequencing studies of human tissue.

We used enricher function from clusterProfiler to perform enrichment analysis.

```{r fig.width=10, fig.height=10, out.width='100%'}
# define background genes = universe for enrichment
background <-  row_metadata$gene_ids

## define genesets
c8 <- msigdbr(species = "human", category = "C8")

msigdbr_hallmarks = hallmarks %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
msigdbr_c3 = c3 %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
msigdbr_c8 = c8 %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
```

#### Hallmarks MSiGDB gene sets

```{r fig.width=25, fig.height=10, out.width='100%'}
tmp_H <- NULL
for(i in unique(srat$seurat_clusters)){
signature = s.markers$feature[s.markers$group1 == i]
ego_module <- enricher(gene = signature, universe = background, TERM2GENE = msigdbr_hallmarks)
tmp <- NULL

  if(!is.null(ego_module)){
    if(dim(ego_module)[1]>0){
    b_H <- barplot(ego_module, showCategory = 15, font.size = 20, label_format = 40)  
    tmp <- b_H$data
    tmp$set <- "Hallmarks"
    tmp$cluster <- i
    }
  }
tmp_H <- rbind(tmp_H, tmp)
}



ggplot(tmp_H, aes(Count, ID)) +
    geom_bar(stat = "identity", aes(fill=cluster))+

  theme_classic()+ 
    geom_text(
        aes(label = (paste( "Padj = ", round(p.adjust,2)))),
        color = "black",
        size = 3,
        hjust=1,
        position = position_dodge(0.5)
    )+
   theme(text = element_text(size=14))+
  facet_wrap(facets=c("cluster"), ncol = length(unique(tmp_H$cluster)))
```


EMT signature should be enriched in stroma cluster.
E2F/proliferation should be enriched in blastema cluster.
MYC(N), TP53 must be enriched in blastema cluster.

#### C3 MSiGDB regulatory target gene sets 

```{r fig.width=25, fig.height=10, out.width='100%'}

tmp_H <- NULL
for(i in unique(srat$seurat_clusters)){
signature = s.markers$feature[s.markers$group1 == i]
ego_module <- enricher(gene = signature, universe = background, TERM2GENE = msigdbr_c3)
tmp <- NULL

  if(!is.null(ego_module)){
    if(dim(ego_module)[1]>0){
    b_H <- barplot(ego_module, showCategory = 15, font.size = 20, label_format = 40)  
    tmp <- b_H$data
    tmp$ID <- factor(tmp$ID, levels = c(tmp$ID[order(tmp$Count)]))
    tmp$set <- "c8"
    tmp$cluster <- i
    }
  }
tmp_H <- rbind(tmp_H, tmp[1:5,])
}



ggplot(tmp_H, aes(Count, ID)) +
    geom_bar(stat = "identity", aes(fill=cluster))+

  theme_classic()+ 
    geom_text(
        aes(label = (paste( "Padj = ", round(p.adjust,2)))),
        color = "black",
        size = 3,
        hjust=1,
        position = position_dodge(0.5)
    )+
   theme(text = element_text(size=14))+
  facet_wrap(facets=c("cluster"), ncol = length(unique(tmp_H$cluster)))
```

Here to check of we catch and MIR pattern (DROSHA; DICER1; other?)

#### C8 MSiGDB cell type signature gene sets

```{r fig.width=30, fig.height=15, out.width='100%'}

tmp_H <- NULL
for(i in unique(srat$seurat_clusters)){
signature = s.markers$feature[s.markers$group1 == i]
ego_module <- enricher(gene = signature, universe = background, TERM2GENE = msigdbr_c8)
tmp <- NULL

  if(!is.null(ego_module)){
    if(dim(ego_module)[1]>0){
    b_H <- barplot(ego_module, showCategory = 15, font.size = 20, label_format = 40)  
    tmp <- b_H$data
    tmp$ID <- factor(tmp$ID, levels = c(tmp$ID[order(tmp$Count)]))
    tmp$set <- "c8"
    tmp$cluster <- i
    }
  }
tmp_H <- rbind(tmp_H, tmp[1:5,])
}



ggplot(tmp_H, aes(Count, ID)) +
    geom_bar(stat = "identity", aes(fill=cluster))+

  theme_classic()+ 
    geom_text(
        aes(label = (paste( "Padj = ", round(p.adjust,2)))),
        color = "black",
        size = 3,
        hjust=1,
        position = position_dodge(0.5)
    )+
   theme(text = element_text(size=14))+
  facet_wrap(facets=c("cluster"), ncol = length(unique(tmp_H$cluster)))
```

The MSigDB C8 gene set is quite relevant for kidney and nephroblastoma annotations. 
Epithelial (cancer and normal) cells should be enriched in mature/adulte kidney pathways while blastema cancer cells will show enrichment of fetal kidney development pathway / cap mesenchyme. 


### Annotation from SingleR (no cancer or kidney specific dataset)

Here, we quickly checked annotations that are present in the _processed rds object. 
However, the automated annotation have not been performed using a cancer specific reference or a kidney reference. 
We do not expect a nice labelling of the cells as the overlap of cell types between the reference and the query dataset is poor. 
This support the need to do a proper label transfer from the fetal kidney atlas, which is imho the best reference that can be applied to a Wilms tumor query. 

```{r fig.width=15, fig.height=10, out.width='100%'}

d2 <-DimPlot(srat, group.by = "singler_celltype_annotation", reduction = "umap", label = TRUE, repel = TRUE) + NoLegend()



DT::datatable(table(srat$seurat_clusters, srat$singler_celltype_annotation), caption = ("table of SingleR annotation per seurat clusters"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

                               buttons = c( 'csv', 'excel')))

d3 <- SCpubr::do_BarPlot(sample = srat,
                         group.by = "singler_celltype_annotation",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file)

d2|d3
```


### Annotation from cellassign (no cancer or kidney specific dataset)

```{r fig.width=15, fig.height=10, out.width='100%'}



d2 <-DimPlot(srat, group.by = "cellassign_celltype_annotation", reduction = "umap", label = TRUE, repel = TRUE) + NoLegend()


DT::datatable(table(srat$seurat_clusters, srat$cellassign_celltype_annotation)
, caption = ("table of cellassign annotation per seurat clusters"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

                               buttons = c( 'csv', 'excel')))

d3 <- SCpubr::do_BarPlot(sample = srat,
                         group.by = "cellassign_celltype_annotation",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file)

d2|d3
```

Even if the result of SingleR and CellAssign are not specific to a Wilms tumor dataset, it give the first impression that C12 is a cluster of endothelial cells and C11 a cluster of immune cells. 

### Azimuth annotation from fetal kidney

To be added in a next PR.


For more information related to the reference, please go to https://www.kidneycellatlas.org/
You will find:

- interactive viewer

- h5ad files to download.

Please note that as Wilms tumor have been described to be closer to fetal kidney as mature kidney, we only used the fetal kidney atlas as the reference. 
Also check : https://www.science.org/doi/10.1126/science.aat5031

This part is imho one of the most important step that allow us to have a quick and reliable idea of the composition of the different clusters. 
The predicted compartment are defined into 4 categories:

- endothelium
- stroma
- fetal nephron 
- immune

As for SingleR and CellAssign, the annotation of immune cells and endothelial cells is straightforward. 
The stroma compartment should then contain normal and cancer stromal cells.
The fetal nephron compartment contain blastema cancer cells a well as normal and cancer epithelial cells. 

Further segregation of cancer versus normal cells will be achieved using a combination of markers/pathways (see above) and inferred CNV (to be done).





## Next step

-[x] We will adapt this template to be render on all the 40 Wilms tumor samples of the dataset. 

-[ ] We will save for each sample the rds file

-[ ] We will run inferCNV for each sample to decide the malignant/normal status of some stroma and epithelial cluster, and confirm the blastema annotation. 

The next step will provide us a better understanding of the entire cohort. We will then have to set up a strategy to annotate each sample. Open questions are:


[ ] should we annotate single cell                      
or   
[] consider applying similar annotations to all cells in a cluster?


[ ] manual annotation of each cluster / each patient    
or   
[] automated annotation using some threshold?

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```
