---
title: "`SCPCL000824` - Using `SingleR` for classification of tumor cells"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
---

## Introduction

This notebook looks at using `SingleR` to annotate tumor cells in `SCPCL000824`. 
We used three different references with `SingleR`: 

1. Both the full `BlueprintEncodeData` reference from `celldex` and `SCPCL000822` where cells are either annotated as tumor or normal. 
Ambiguous cells are removed from `SCPCL000822`. 
2. `SCPCL000822` where all tumor cells are labeled as tumor and all other cells keep the original annotation obtained from running `SingleR` as part of `scpca-nf`.  
3. `SCPCL000822` with tumor cells labeled as in #2 combined with the full `BlueprintEncodeData` reference. 


## Setup

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_bw()
)

# quiet messages
options(readr.show_col_types = FALSE)
ComplexHeatmap::ht_opt(message = FALSE)
```


```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "2024-05-01")
sample_dir <- file.path(data_dir, "SCPCP000015", "SCPCS000492")

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-ewings")
```

```{r}
# source in helper functions: plot_gene_heatmap() and plot_cnv_heatmap() 
# create_classification_df() and create_marker_gene_df()
validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
source(validation_functions)

# source in helper functions for make_jaccard_matrix() and jaccard()
jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
source(jaccard_functions)
```

```{r}
# Input files
sce_file <- file.path(sample_dir, "SCPCL000824_processed.rds")
marker_genes_file <- file.path(module_base, "references", "tumor-marker-genes.tsv")

# annotations from SCPCL000824_tumor-gene-expression-exploration.Rmd
annotations_dir <- file.path(module_base, "results", "annotation_tables", "SCPCS000492")
annotations_file <- file.path(annotations_dir, "SCPCL000824_tumor-classifications.tsv.gz")

# gene set scores to use for validation 
results_dir <- file.path(module_base, "results", "annotate_tumor_cells_output", "SCPCS000492")
geneset_scores_file <- file.path(results_dir, "SCPCL000824_gene-set-scores.tsv")

# singler results 
full_singler_results <- file.path(module_base, "scratch", "SingleR", "SCPCL000824_singler-results.rds")
singler_annotations_file <- file.path(annotations_dir, "SCPCL000824_singler-classifications.tsv.gz")
```

```{r}
# read in sce file 
sce <- readr::read_rds(sce_file)

# read in annotations
annotations_df <- readr::read_tsv(annotations_file)
singler_annotations_df <- readr::read_tsv(singler_annotations_file)

# read in full singler results 
singler_results_list <- readr::read_rds(full_singler_results)

# read in gene set scores
geneset_scores_df <- readr::read_tsv(geneset_scores_file)
```


```{r}
# generate classification df to use for plots 
classification_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
  ) |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2, singler_celltype_annotation) |> 
  # join with previous annotations, singler results, and gene set scores 
  dplyr::left_join(annotations_df, by = c("barcodes" = "cell_barcode")) |> 
  dplyr::left_join(singler_annotations_df, by = "barcodes") |> 
  dplyr::left_join(geneset_scores_df, by = "barcodes") |>
  # get the top cell types for each classification method for plotting later 
  dplyr::mutate(
    across(
      ends_with("_classification"),
      \(x) {
        x |>
          forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
          forcats::fct_infreq() |>
          forcats::fct_relevel("All remaining cell types", after = Inf)
      },
      .names = "{.col}_lumped"
    )
  )
```

## SingleR results 

### UMAPs of cell assignments 

The first thing we will do is look at the annotated cell types on a UMAP. 
The top 7 cell types are shown for each reference used and all other cells are lumped together. 

```{r}
# list of columns with singler results for UMAPs
singler_annotation_columns <- c(
  "blueprint_tumor_classification_lumped",
  "updated_singler_classification_lumped",
  "updated_singler_blueprint_classification_lumped"
) |> 
  purrr::set_names("Blueprint", "Original", "Original + Blueprint")

singler_annotation_columns |> 
  purrr::map(\(annotation_column){
    ggplot(
    classification_df,
    aes(x = UMAP1, y = UMAP2, color = .data[[annotation_column]])
  ) +
    # set points for all "other" points
    geom_point(
      data = dplyr::select(
        classification_df, -.data[[annotation_column]]
      ),
      color = "gray80",
      alpha = 0.5,
      size = 0.5
    ) +
    # set points for desired cell type
    geom_point(size = 0.5, alpha = 0.5) +
    facet_wrap(
      vars(.data[[annotation_column]]),
      ncol = 3
    ) +
    scale_color_brewer(palette = "Dark2") +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    guides(
      color = guide_legend(
        title = "Cell types",
        # more visible points in legend
        override.aes = list(
          alpha = 1,
          size = 1.5
        )
      )
    )
  })
```

It appears that most cells in that bottom group of cells are being called tumor cells regardless of the reference that is used. 
The major differences here are in how the normal cells are labeled. 
For right now, let's focus on just identifying the tumor cells and see how similar each of the references are to each other before we compare to the other methods that we have used. 

### Compare tumor cell annotations between references used 

First we will just look at the overlap in the tumor cells between all three methods. 

```{r}
tumor_cells <- list(
  "Blueprint" = classification_df$barcodes[classification_df$blueprint_tumor_classification == "Tumor"],
  "Original" = classification_df$barcodes[classification_df$updated_singler_classification == "Tumor"],
  "Original and Blueprint" = classification_df$barcodes[classification_df$updated_singler_blueprint_classification == "Tumor"]
)

UpSetR::upset(UpSetR::fromList(tumor_cells), order.by = "freq") |> print()
```

Let's see where the shared cells are and if they are where we would expect them to be based on marker gene expression plots we made previously (bottom right group of cells). 

```{r}
classification_df <- classification_df |> 
  dplyr::mutate(singler_consensus_call = dplyr::case_when(
    blueprint_tumor_classification == "Tumor" & updated_singler_classification == "Tumor" & updated_singler_blueprint_classification == "Tumor" ~ "Tumor",
    blueprint_tumor_classification == "Tumor" | updated_singler_classification == "Tumor" | updated_singler_blueprint_classification == "Tumor" ~ "Ambiguous",
    .default = "Normal"
  )) 

ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = singler_consensus_call)) +
  geom_point(alpha = 0.5, size = 0.5)
```

It looks like the cells that are normal are in the smaller clusters and the larger cluster is mostly tumor cells with some that are ambiguous across `SingleR` references. 

### Compare normal cell annotations between references used 

Next we will compare all of the top cell types identified in each method to the top cell types identified in the other methods. 
This will show us the overlap in tumor cells and normal cell types identified. 

```{r, fig.height=7}
methods <- c(
  "updated_singler_classification_lumped",
  "updated_singler_blueprint_classification_lumped"
) |> 
  purrr::set_names("Original", "Original and Blueprint")

# compare blueprint to original and origintal + blueprint
plot_jaccard(classification_df,
             annotation_column = "blueprint_tumor_classification_lumped",
             methods_to_compare = methods,
             column_title = "Blueprint",
             legend_match = "Original") |> 
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )

# compare original to original + blueprint
plot_jaccard(classification_df,
             annotation_column = "updated_singler_classification_lumped",
             methods_to_compare = c("Original and Blueprint" = "updated_singler_blueprint_classification_lumped"),
             column_title = "Original",
             legend_match = "Original")
  
```

- There seems to be mostly consensus across all three methods on which cells are tumor cells. 
- When only using Blueprint, a lot of cells are categorized as "Normal". 
This means they match the non-tumor cells identified in `SCPCL000822` more than they match the `Blueprint` reference. 
Additionally, with only Blueprint, there is a group of cells that are called as tumor but actually appear to match endothelial cells present in `SCPCL000822`.
- The cells that aren't called as either Tumor or Normal in the Blueprint only reference mostly match up to only `erythrocyte` in the reference that just uses `SCPCL000822`. 
These same cells match to their corresponding Blueprint cell type when combining `SCPCL000822` and Blueprint (e.g., fibroblasts, neurons, and epithelial cells are identifed in both). 
- When comparing using just `SCPCL000822` to `SCPCL000822` and Blueprint, we see a overlap in endothelial cells and chrondrocytes. 
We also see that for the most part the top cell types are the same except when using blueprint you see the addition of epithelial cells. 


Next, we will measure how well the updated classification matches the original `SingleR` class identified in `SCPCL000824`. 

```{r}
singler_annotation_columns |> 
  purrr::map(\(classification_column){
    
    # first get list of tumor cells for this classificationn
    # we want to remove these since we are only interested in comparing normal cell types
    tumor_cells <- classification_df |> 
      dplyr::filter({{classification_column}} == "Tumor") |> 
      dplyr::pull(barcodes)
    
    # get list of shared cell types with SingleR
    original_celltypes <- caret_df$singler_celltype_annotation |> 
      unique()
    new_celltypes <- caret_df[[classification_column]] |> 
      unique()
    
    shared_celltypes <- intersect(original_celltypes, new_celltypes)
    
    # don't filter, just label as tumor? or other normal?  
    caret_df <- caret_df |> 
      # first remove tumor cells 
      dplyr::filter(!barcodes %in% tumor_cells) |> 
      # then update to make sure all cell types are shared
      # anything not in shared cell types are classified as "other" 
      dplyr::mutate(singler_caret = dplyr::if_else(singler_celltype_annotation %in% shared_celltypes, singler_celltype_annotation, "Other"),
                    new_ref_caret = dplyr::if_else(!!sym(classification_column) %in% shared_celltypes, !!sym(classification_column), "Other"))
    
    caret::confusionMatrix(
      table(
        caret_df$singler_caret, 
        caret_df$new_ref_caret)
    )
    
  })
```

Here we see that the highest Kappa, although still pretty low, is with using the original `SingleR` annotations from `SCPCL000822` and use of Blueprint. 

### Compare tumor cell annotations with `SingleR` to marker gene based methods 


```{r fig.height=7}
 original_methods <- c(
    "updated_marker_gene_classification",
    "auc_classification",
    "ucell_classification"
  ) |> 
  purrr::set_names("Marker genes", "AUCell", "UCell")

# heatmap comparing each singler annotation type to each of the marker gene based methods 
heatmap_list <- singler_annotation_columns |> 
  purrr::iwalk(\(annotation_column, name) {
    
    # first make singler annotations normal or tumor for easier visualization
    plot_df <- classification_df |> 
      dplyr::mutate(new_annotation = dplyr::if_else(!!sym(annotation_column) == "Tumor", "Tumor", "Normal"))
    
    plot_jaccard(
      classification_df = plot_df, 
      annotation_column = "new_annotation", 
      methods_to_compare = original_methods,
      column_title = name,
      legend_match = "UCell"
    ) |> 
      ComplexHeatmap::draw(
        heatmap_legend_side = "right",
        # add a margin to the heatmap so labels don't get cut off
        padding = unit(c(2, 20, 2, 2), "mm")
      )
  })
```

Generally, these methods agree on which cells are tumor cells. 

It looks like using Blueprint alone is the most similar to all marker based methods. 
However, it looks like `UCell` categorizes more cells as normal that are classified as tumor with any of the `SingleR` methods. 

For both original and original and blueprint, it looks like there is a good amount of cells that `SingleR` classifies as normal, but the marker gene based methods classify as tumor. 

### Validate marker gene expression in `SingleR` annotations 

The last thing we will do is confirm that the cells that are identified as tumor cells with `SingleR` have increased expression of marker genes and higher gene set scores. 
We would expect this to be the case, but we also want to see if we are missing calling any tumor cells that have higher gene expression. 

For these plots we will also look at all the marker based classifications so we can compare across all options. 

```{r}
# first get marker gene expression 
plot_markers_df <- create_marker_gene_df(
  sce,
  classification_df, 
  marker_genes_file
)
```

```{r}
# create a density plot showing the distribution of marker gene expression across classification methods 
marker_density_df <- plot_markers_df |>
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = dplyr::case_when(
      method == "marker_gene_classification" ~ "Marker genes only",
      method == "updated_marker_gene_classification" ~ "Marker gene cutoff from 822",
      method == "auc_classification" ~ "AUCell",
      method == "ucell_classification" ~ "UCell",
      method == "blueprint_tumor_classification" ~ "Blueprint",
      method == "updated_singler_classification" ~ "Original SingleR with tumor",
      method == "updated_singler_blueprint_classification" ~ "Tumor + SingleR + Blueprint"
    ),
    classification = dplyr::if_else(classification == "Tumor", "Tumor", "Other")
  )

ggplot(marker_density_df, aes(x = sum_transformed_exp, color = classification)) +
  geom_density() +
  facet_wrap(vars(method)) +
  labs(
    title = "Marker gene expression",
    x = "Sum of marker genes"
  )
```


```{r fig.height=10}
# plot gene set scores for each cell  
geneset_plot_df <- classification_df |> 
  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("mean-")) |> 
  tidyr::pivot_longer(
    cols = starts_with("mean"),
    names_to = "geneset",
    values_to = "mean_score"
  ) |> 
  dplyr::mutate(
    geneset = stringr::word(geneset, -1, sep = "-")
  ) |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = dplyr::case_when(
      method == "marker_gene_classification" ~ "Marker genes only",
      method == "updated_marker_gene_classification" ~ "Marker gene cutoff from 822",
      method == "auc_classification" ~ "AUCell",
      method == "ucell_classification" ~ "UCell",
      method == "blueprint_tumor_classification" ~ "Blueprint",
      method == "updated_singler_classification" ~ "Original SingleR with tumor",
      method == "updated_singler_blueprint_classification" ~ "Tumor + SingleR + Blueprint"
    ),
    classification = dplyr::if_else(classification == "Tumor", "Tumor", "Other")
    ) 
  
ggplot(geneset_plot_df, aes(x = mean_score, color = classification)) +
  geom_density(bw = 0.05) +
  facet_grid(rows = vars(method), 
             cols = vars(geneset)) +
  labs(
    title = "Gene set scores",
    x = "Mean gene set score"
  )

```

Looking at these plots, generally Blueprint only has the best separation between tumor cells and all other cells for both gene set scores and marker gene expression. 
This also seems to mirror the plots shown with `AUCell` as the classifier. 

It looks like we may be mis-classifying some cells when using the original `SingleR` assignments for SCPCL000822 with or without Blueprint. 
There is some overlap in the cells that are identified as tumor and other non-tumor cells. 

## Conclusions 

## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```
