---
title: "Summary of tumor cell classification with `SingleR` for `r params$library_id`"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
params:
  library_id: NULL
  sce_file: NULL
  singler_results_file: NULL
  geneset_scores_file: NULL
  marker_genes_file: NULL
---

This report summarizes the classification of `r params$library_id` using the following approach: 

- Tumor cells were identified by using `AUCell`. 
- All cells were labeled using `SingleR` with a combination of three references: 
  - All tumor cells annotated by `AUCell` for all other libraries in `SCPCP000015`. 
  - `BlueprintEncodeData` from `celldex`
  - `HumanPrimaryCellAtlasData` from `celldex`


## Setup

```{r}
# check that sce and results files exist
stopifnot(
  "sce file does not exist" = file.exists(params$sce_file),
  "SingleR results file does not exist" = file.exists(params$singler_results_file),
  "Gene set scores file does not exist" = file.exists(params$geneset_scores_file),
  "Marker genes file does not exist" = file.exists(params$marker_genes_file)
)
```


```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_bw()
)

# quiet messages
options(readr.show_col_types = FALSE)
ComplexHeatmap::ht_opt(message = FALSE)
```


```{r base paths}
# The path to this module
module_base <- rprojroot::find_root(rprojroot::is_renv_project)

# source in helper functions: plot_gene_heatmap() and plot_cnv_heatmap()
# create_classification_df() and create_marker_gene_df()
validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
source(validation_functions)
```

```{r}
# Read in files
sce <- readr::read_rds(params$sce_file)
singler_results_df <- readr::read_tsv(params$singler_results_file)
geneset_scores_df <- readr::read_tsv(params$geneset_scores_file)
```

## Prep for plots

```{r}
# generate classification df to use for plots
classification_df <- sce |>
  scuttle::makePerCellDF(use.dimred = "UMAP") |>
  # replace UMAP.1 with UMAP1 and get rid of excess columns
  dplyr::select(barcodes, UMAP1 = UMAP.1, UMAP2 = UMAP.2) |>
  # join with previous annotations, singler results, and gene set scores
  dplyr::left_join(singler_results_df, by = "barcodes") |>
  dplyr::left_join(geneset_scores_df, by = "barcodes") |>
  dplyr::mutate(
    # set factor order for aucell
    aucell_annotation = forcats::fct_relevel(aucell_annotation, "tumor", after = 0),
    # first remove other library IDs from tumor annotations
    # we only care about whether or not it was from the original library or a new library
    singler_annotation = dplyr::case_when(
      stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
      is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
      .default = singler_annotation
    ) |>
      forcats::fct_relevel("tumor", after = 0),
    # get the top cell types for plotting later
    singler_lumped = singler_annotation |>
      # there are warnings because we have some that have less than 7 cell types (ones that only did tumor/normal)
      # this is why we ignore warnings in this chunk
      forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
      forcats::fct_infreq() |>
      forcats::fct_relevel("All remaining cell types", after = Inf),
    # make a label that is just tumor or normal
    singler_tumor_normal = dplyr::if_else(!(singler_annotation %in% c("tumor", "unknown")), "normal", singler_annotation) |>
      forcats::fct_relevel("tumor", after = 0),
    # get the consensus between aucell and singler
    consensus = dplyr::case_when(
      singler_annotation == "tumor" & aucell_annotation == "tumor" ~ "tumor_both",
      singler_annotation == "tumor" & aucell_annotation != "tumor" ~ "tumor_singler",
      singler_annotation != "tumor" & aucell_annotation == "tumor" ~ "tumor_aucell",
      .default = "normal"
    ) |>
      forcats::fct_relevel("normal", after = Inf)
  )
```


```{r}
# read in marker genes table
marker_genes_df <- readr::read_tsv(params$marker_genes_file, show_col_types = FALSE) |>
  # account for genes being from multiple sources
  dplyr::select(cell_type, ensembl_gene_id, gene_symbol) |>
  dplyr::distinct()

# get list of all cell types found
cell_types <- unique(marker_genes_df$cell_type)

# get the sum of expression of all genes for each cell type
gene_exp_df <- cell_types |>
  purrr::map(\(type){
    calculate_sum_markers(marker_genes_df, sce, type)
  }) |>
  purrr::reduce(dplyr::inner_join, by = "barcodes")

# join sum expression columns with classification df
classification_df <- classification_df |>
  dplyr::left_join(gene_exp_df, by = "barcodes")
```

```{r}
# get columns that correspond to marker genes and gene sets to help with plotting later
marker_gene_columns <- colnames(classification_df)[which(endsWith(colnames(classification_df), "_sum"))]
geneset_columns <- colnames(classification_df)[which(startsWith(colnames(classification_df), "mean-"))]
```

## UMAPs of all cell type annotations 

Here we show the annotated cell types on a UMAP. 
The top 7 cell types are shown for each reference used and all other cells are lumped together. 

```{r, fig.height=8, fig.width=8}
plot_faceted_umap(classification_df, singler_lumped)
```


## Comparison between `SingleR` and `AUCell`

Below we show the cells that are annotated as tumor by both methods, just `AUCell` and just `SingleR` along with all normal cells. 

```{r}
plot_faceted_umap(classification_df, consensus)
```

We then calculate the confusion matrix between tumor and normal cells called in `SingleR` and `AUCell`. 
The rows here correspond to the `AUCell` annotation and columns are from `SingleR`. 

```{r}
caret_df <- classification_df |>
  dplyr::filter(singler_tumor_normal != "unknown") |>
  dplyr::mutate(singler_tumor_normal = droplevels(singler_tumor_normal, "unknown"))

# get the cell types in both
# these should be tumor normal but sometimes that's not the case
singler_cell_types <- unique(caret_df$singler_tumor_normal) |> as.character()
aucell_cell_types <- unique(caret_df$aucell_annotation) |> as.character()

# only calculate confusion matrix if all cell types are the same
if (all(singler_cell_types == aucell_cell_types)) {
  caret::confusionMatrix(
    table(
      caret_df$aucell_annotation,
      caret_df$singler_tumor_normal
    )
  )
}
```


## Validate output from `SingleR`

The next three sections will evaluate the annotations output obtained from using `SingleR` by looking at expression of marker genes and gene set scores. 

The gene expression for all marker genes for tumor cells, endothelial cells, mesenchymal-like cells, and immune cells was summed together for each cell. 
Marker genes were obtained from [Visser et al.](https://doi.org/10.1158/2767-9764.CRC-23-0027). 

Gene set scores are determined by calculating the mean expression of all genes in a given gene set. 
Scores are calculated for the following gene sets from `MsigDb`:

  - [`ZHANG_TARGETS_OF_EWSR1_FLI1_FUSION`](https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/ZHANG_TARGETS_OF_EWSR1_FLI1_FUSION.html)
  - [`RIGGI_EWING_SARCOMA_PROGENITOR_UP`](https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/RIGGI_EWING_SARCOMA_PROGENITOR_UP.html?ex=1)
  - [`SILIGAN_TARGETS_OF_EWS_FLI1_FUSION_DN`](https://www.gsea-msigdb.org/gsea/msigdb/cards/SILIGAN_TARGETS_OF_EWS_FLI1_FUSION_DN)

For both marker gene expression and gene set scores, the following plots will be shown: 

- Heatmap showing gene expression or gene set scores across all cells. 
Here the rows correspond to the gene list and the columns are the cells. 
The annotations are shown below the heatmap. 
- Density plot showing gene expression or gene set scores across all cells. 
Each row is a cell type and the expression or score is plotted on the x-axis. 

```{r}
# check that marker genes are expressed, otherwise turn off those plots
total_exp <- sum(classification_df[marker_gene_columns])
if(total_exp > 0){
  show_marker_gene_plots <- TRUE
} else {
  show_marker_gene_plots <- FALSE
  message("No expression of marker genes in this library. No plots will be displayed in sections labeled 'Marker gene expression'.")
}

```

```{r}
# check that gene set scores aren't all 0, otherwise turn off those plots
total_score <- sum(classification_df[geneset_columns])
if(total_score > 0){
  show_gene_set_plots <- TRUE
} else {
  show_gene_set_plots <- FALSE
  message("Genes present in provided gene sets are not expressed in this library. No plots will be displayed in sections labeled 'Gene set scores'.")
}

```


### Tumor vs. Normal 

In this section we show just the cells that are considered tumor and normal, lumping all non-tumor cell types together. 

**Marker gene expression**


```{r, eval=show_marker_gene_plots}
full_celltype_heatmap(classification_df, marker_gene_columns, "singler_tumor_normal") 
```


```{r, eval=show_marker_gene_plots}
plot_density(
  classification_df,
  "tumor_sum",
  "singler_tumor_normal"
)
```

**Gene set scores**

```{r, eval=show_gene_set_plots}
full_celltype_heatmap(classification_df, geneset_columns, "singler_tumor_normal")
```

```{r, fig.height=10, eval=show_gene_set_plots}
geneset_columns |>
  purrr::map(\(column){
    plot_density(
      classification_df,
      column,
      "singler_tumor_normal"
    )
  }) |>
  patchwork::wrap_plots(ncol = 1)
```


### Tumor vs. all other cell types 

In this section we show all tumor cells and the top 5 most represented normal cell types.

**Marker gene expression** 

```{r, eval=show_marker_gene_plots}
full_celltype_heatmap(classification_df, marker_gene_columns, "singler_lumped")
```

```{r, fig.height=10, eval=show_marker_gene_plots}
marker_gene_columns |>
  purrr::map(\(column){
    plot_density(
      classification_df,
      column,
      "singler_lumped"
    )
  }) |>
  patchwork::wrap_plots(ncol = 1)
```


**Gene set scores**

```{r, eval=show_gene_set_plots}
full_celltype_heatmap(classification_df, geneset_columns, "singler_lumped")
```

```{r, fig.height=10, eval=show_gene_set_plots}
geneset_columns |>
  purrr::map(\(column){
    plot_density(
      classification_df,
      column,
      "singler_lumped"
    )
  }) |>
  patchwork::wrap_plots(ncol = 1)
```

### Tumor cells annotated by `SingleR` and `AUCell`

Here we compare the marker gene expression and gene set scores for cells annotated as tumor by both methods, just `SingleR`, or just `AUCell` and compare to all normal cells. 

**Marker gene expression**

```{r, eval=show_marker_gene_plots}
full_celltype_heatmap(classification_df, marker_gene_columns, "consensus")
```


```{r, eval=show_marker_gene_plots}
plot_density(
  classification_df,
  "tumor_sum",
  "consensus"
)
```


**Gene set scores**

```{r, eval=show_gene_set_plots}
full_celltype_heatmap(classification_df, geneset_columns, "consensus")
```

```{r, fig.height=10, eval=show_gene_set_plots}
geneset_columns |>
  purrr::map(\(column){
    plot_density(
      classification_df,
      column,
      "consensus"
    )
  }) |>
  patchwork::wrap_plots(ncol = 1)
```


## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```


