---
title: "Explore normal cell types in Neuroblastoma samples"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
---


## Introduction

The goal of this notebook is to explore the distribution of normal consensus cell types, specifically immune and endothelial, in the Neuroblastoma tumor (i.e., _not_ the PDX) samples from `SCPCP000004`.

Based on these findings, we can determine approaches to evaluate for pooling cells to use as a normal reference for `inferCNV`.
We can also determine whether any samples can be run with an internal (non-pooled) reference as a control.

## Setup

```{r}
options(readr.show_col_types = FALSE)
library(ggplot2)
theme_set(theme_bw())
```

### Paths

#### Base directories


```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current", "SCPCP000004")

# The current results directory, found within the repository base directory
result_dir <- file.path(repository_base, "data", "current", "results")

# Path to cell-type-consensus results
consensus_dir <- file.path(result_dir, "cell-type-consensus", "SCPCP000004")

# Path to reference cell type groupings
reference_dir <- file.path(repository_base, "analyses", "infercnv-consensus-cell-type", "references")
```

#### Input and output files

```{r paths}
# Input consensus cell type files
celltype_files <- list.files(
  path = consensus_dir,
  pattern = "_processed_consensus-cell-types\\.tsv\\.gz$",
  recursive = TRUE,
  full.names = TRUE
) |>
  # set names as library id
  purrr::set_names(
    \(x) {
      stringr::str_split_i(basename(x), "_", 1)
    }
  )

# project sample metadata file
metadata_file <- file.path(data_dir, "single_cell_metadata.tsv")

# reference cell type groups
reference_file <- file.path(reference_dir, "reference-cell-groups.tsv")
```


## Analysis

```{r}
# determine PDX libraries to exclude
tumor_library_ids <- readr::read_tsv(metadata_file) |>
  dplyr::filter(!is_xenograft) |>
  dplyr::pull(scpca_library_id)

# get cell groups
cell_group_df <- readr::read_tsv(reference_file)

# Prepare data frame of all cell types
celltype_df <- celltype_files |>
  # consider only non-PDX
  purrr::keep_at(tumor_library_ids) |>
  purrr::map(readr::read_tsv) |>
  purrr::list_rbind() |>
  dplyr::select(library_id, barcodes, consensus_annotation) |>
  dplyr::left_join(cell_group_df, by = "consensus_annotation")
```

We'll first exclude any libraries that have fewer than 1000 cells from consideration in the reference.

```{r}
discard_libraries <- celltype_df |>
  dplyr::count(library_id) |>
  dplyr::filter(n < 1000) |>
  dplyr::pull(library_id)

discard_libraries
```

```{r}
celltype_df <- celltype_df |>
  dplyr::filter(!(library_id %in% discard_libraries))
```

These cell groupings are present in the neuroblastoma samples we're considering:

```{r}
unique(celltype_df$reference_cell_group)
```

We should remove any cell groups which might correspond to tumor cells, including `stromal`, `neuronal`, and `stem cell`.
The `NA` values here correspond to `Unknown` cells, which we'll also remove.
This means we'll keep groups `endothelial`, `epithelial`, and `immune`.

```{r}
exclude_groups <- c("stromal", "neuronal", "stem cell")
celltype_df <- celltype_df |>
  dplyr::filter(
    !(reference_cell_group %in% exclude_groups),
    consensus_annotation != "Unknown"
  )
```



Let's begin by counting how many cells in each cell group we have per library

```{r}
celltype_count_df <- celltype_df |>
  dplyr::count(library_id, reference_cell_group)

# format for viewing
celltype_count_wide_df <- celltype_count_df |>
  tidyr::pivot_wider(
    names_from = reference_cell_group,
    values_from = n,
    values_fill = 0
  ) |>
  # add column for total normal cells
  dplyr::mutate(
    all_normal = rowSums(dplyr::across(where(is.numeric)))
  ) |>
  # arrange by total normal cells
  dplyr::arrange(desc(all_normal))

# this way we can print the full table at once; RMarkdown will only show 10 rows at once
knitr::kable(celltype_count_wide_df)
```

Virtually all of the normal cell types are `endothelial` and `immune`, with very few `adipocyte` and literally only one `epithelial` cell.

We'll look at this data in barplot form to see, per-library, the relative proportions of each normal cell type group out of all normal cells in the library.


```{r}
celltype_prop_df <- celltype_count_wide_df |>
  # long again, but keep all_normal column for calculating proportion
  tidyr::pivot_longer(
    -c(library_id, all_normal),
    names_to = "reference_cell_group",
    values_to = "n"
  ) |>
  dplyr::mutate(cell_group_prop = n / all_normal)

ggplot(celltype_prop_df) +
  aes(x = library_id, y = cell_group_prop, fill = reference_cell_group) +
  geom_col() +
  labs(y = "Proportion of normal cells in group") +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 30, hjust = 1, size = 7)
  )
```

Based on these numbers, we can consider three normal reference options:

* All normal cells
* ALl immune cells
* All endothelial cells

How many cells would be in each of those pooled references?

```{r}
# update the count data frame without categories we don't need
celltype_count_df <- celltype_count_wide_df |>
  dplyr::select(-adipocyte, -epithelial) |>
  tidyr::pivot_longer(
    -library_id,
    names_to = "reference_cell_group",
    values_to = "n"
  )

celltype_count_df |>
  dplyr::group_by(reference_cell_group) |>
  dplyr::summarize(total_cells = sum(n))
```

Well, that's quite a lot of cells for these pooled references!
A potential concern is that this many cells will incur prohibitive runtimes, so we may want to explore a subsampling strategy.
Any subsampling strategy should aim to retain the relative cell type proportions across libraries, but we can explore that if runtime becomes a problem when we reach that stage.

Finally, let's see which libraries could have an internal reference.
`inferCNV` recommends at least 100 cells in a reference, so which combinations of library/reference are possible here?

```{r}
celltype_count_df |>
  dplyr::filter(n >= 100) |>
  dplyr::group_by(library_id) |>
  dplyr::mutate(reference_options = paste(reference_cell_group, collapse = ", ")) |>
  dplyr::ungroup() |>
  dplyr::select(library_id, reference_options) |>
  unique() |>
  # add row counter so we can easily see how many libraries
  dplyr::mutate(row_num = dplyr::row_number()) |>
  # kable to print the full table at once
  knitr::kable()
```

## Conclusions

* To start, we can consider three references: all normal, all immune, and possibly also all endothelial
  * We may need to consider a subsetting strategy for these references if runtimes are too long since there would be >10,000 cells in the immune-containing pooled normal references
* There are 19 libraries we can use internal references with, all with the option for all normal cells and some with the immune-only or endothelial-only options too


## Session Info

```{r session info}
# record the versions of the packages used in this analysis and other environment information
sessionInfo()
```
